Extending the WebID Protocol with Access Delegation

### 2012-07-28 paper.tex
streamline structure + more text

A \textit{WebID} is a URI that refers to an agent - person, robot, group or other thing that can have intentions.
The WebID should be a URI which when dereferenced returns a representation whose description uniquely identifies the agent as the controller of a public key~\cite{sporny-m-2011--a}.

@@

Things get more interesting in the Authorization space.
Consider a very natural application of WebID: allowing friends of one's friends (foaf) access to some resources.
This authorization rule will require the Web Server to fetch each of its users' friends profiles, in order to build up such a list.
But there is a privacy issue involved here: not every one wants to make all of their social network (or even any of it) publicly visible.
Those people may then protect their FOAF Profile with Access Control rules.
How can a server that needs access to these FOAF profiles in order to apply its own access control rules get access to the information?
Should it be able to do so?
These are the questions we will try to answer in this paper.

\section{Solution 1: Using each users WebID}

\section{Solution 2: Acting on Behalf of a User}

\section{Comparison with Other protocols}

\section{OAuth}

\section{CORS}

\cite{story-h-2009--a}
\cite{tramp-s-2012--a}

%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

\section{Terminology}

\textit{Access Delegation} is a relation between two WebID resources.
The relation describes, under which circumstances a secretary agent has access to a certain resource under the same constraints as the named agent.
The secretary agent declares to act in the name of the named agent.

\section{Requirements}

@@

\subsection{Basic Use-Cases for Access Delegation}\label{sec:usecases}

\paragraph{Personal Devices and Applications}

Currently, the default usage intention for WebIDs are browser-based authentication scenarios.
This is an important use-case and should be of special interest but we use the world wide web not only from our browsers but also from other applications and from other application on other devices such as smartphones, tablets and freedomboxes\footnote{\url{https://freedomboxfoundation.org/}}.
In a typical every day usage scenario, we use feed readers and social network clients hand in hand with web browsers.
This is true esp. for mobile devices where specialized clients for social activities are widely in use and does exist even for academically driven social networks as the distributed Social Web \todo{A: req. explaining? (Web2.0 vs 3.0 ambiguity)} %Web2.0 or 3.0? the begining of the sentence talks about 2.0, and you end with FOAF (3.0)?
 based on FOAF WebIDs~\cite{tramp-s-2011--a}.

All these applications have in common that they act in the name of its users and fetch and manipulate resources in a way that this could also be done by the user itself and with her browser.

We distinguish two different classes of applications here.
(1) Applications which provide their service to one user only and run on a device which can be accessed by one user only (e.g. a smartphone).
These applications have more in common with the users browser and can be seen as an extension to the user itself \cite{clark-a-1998-7-a}.

In a WebID-enabled communication scenario, these devices use their own certificates for communication since we do not want to copy the browser certificate to these devices and we also do not want to save the passphrase on theses devices and services.

\todo{Sebastian Tramp: Extend it with respect to the Extended Mind paper}

\paragraph{Vacation Replacement}

In a typical office scenario, employees declare vacation replacements for the time of their holiday in order to achieve an uninterruptible service during that time.
One option to allow access to this vacation replacement is to give him the original certificate or to add his private key to the users WebID.
This solution has two drawbacks: (1) the users certificate is burned after the vacation and she needs to create a new one, (2) a resource guard can not distinguish between the user and the vacation replacement user, which allows for abuse.

In this context, the named agent is the employee which is on holiday and her vacation replacement is the secretary agent.

\paragraph{Working Groups}\label{sec:WorkingGroups}
\todo{A: why 'working' and not just 'Groups'? (difference?)}
Access for working groups to a resource is typically specified in a way that each resource guard \todo{A: is it really the guard who holds the list, or some access control sys?} needs to hold a list of group members of this group.
In the Social Semantic Web \todo{A: see my prev. note about Web2.0 vs 3.0}, these working groups have its own WebID resources which describe the group as an entity and its members as relations to their user WebIDs.
To authorize a group of agents to access a resource should be as simple as to a single agent.
In order to allow this, we define the group as the named agent and all group members as its secretary agents.

An alternative could be to share one single group certificate with all members or to create a group certificate for each member and describe all public keys in the group WebID.
Again, these solutions have different drawbacks: (1) a shared certificate needs to be updated each time a user leaves the group, (2) multiple group certificates for each user does not allow for distinguishing which user accessed the resource, which allows for abuse.

In this context, the group is the named agent and all current group members are secretary agents.\todo{A: not sure I agree (see source)}%the group is not an entity, therefore it cannot take actions by iteself. Actions are always performed by users, as part of the group.

\paragraph{Office Roles}
An office role is a construct where a user can access specific resources because of his working description.
A product manager has access to all product related resources because of her role as product manager and not because of her identity.
The role itself is identified by WebID and all access authorization rules should use this role WebID instead of a user WebID.

In this context, the role is the named agent and the current role owner is the secretary agent.

\subsection{Implementation Requirements}\label{sec:req}

@@

\paragraph{Distinguish secretary agents from named agents}
Delegation should not be handled by camouflaging an request of an agent in the sense that this request is indistinguishable from a named agent request.
A secretary agent should use its own certificate and it should be identified and described by its own WebID profile.
The motivation for this is twofold.
Firstly, it allows resource guards to permit or deny requests based on this information.
Secondly, applications which are secretary for many different named agents do not need to switch their certificate between requests.
Also they do not need to handle hundreds of different keys.

\paragraph{Easy to use}
The one and only place to describe which secretary agents are allowed to operate as a named agent should be the WebID profile itself.
To grant delegated access to a secretary agent, no other actions than simply adding triples to the WebID profile should be needed.
To reset this grant the same triple can removed from the WebID profile.

\paragraph{Linked Data and Read Write Web integration}

@@

\paragraph{Minimal protocol footprint}

@@

\paragraph{No enforcement to reveal agent identities}\todo{maybe we kick this ...}

@@

\section{Specification of the WebID protocol extension}\label{sec:spec}

@@

\subsection{OAuth}
\todo{Andrei Sambra: Compare to oAuth}
\todo{If we were to implement a secreteary like feature using OAuth what would it look like? what would the differences be?}

>>>>

A \textit{WebID} is a URI that refers to an agent - person, robot, group or other thing that can have intentions.
The WebID should be a URI which when dereferenced returns a representation whose description uniquely identifies the agent as the controller of a public key~\cite{sporny-m-2011--a,story-h-2009--a}

@@

Things get more interesting in the authorization space.
Consider a very natural application of WebID: allowing friends of one's friends (foaf) access to some resources.
This authorization rule will require the web server to fetch each of its users' friends profiles, in order to build up such a list.
But there is a privacy issue involved here: not every one wants to make all of their social network (or even any of it) publicly visible.
Those people may then protect their FOAF profile with access control rules.
How can a server that needs access to these FOAF profiles in order to apply its own access control rules get access to the information?
Should it be able to do so?
These are the questions we will try to answer in this paper.

The rest of the paper is organized in the following way:
\autoref{sec:req} describes some more use-cases and additional requirements which we had in mind for our solution,
\autoref{sec:spec} goes into detail with the WebID specification and adds support for authorization delegation,
in \autoref{sec:eval} we describe our reference implementations based on two web applications and finally
in \autoref{sec:relatedwork} we compare our proposal with two protocols, namely CORS and OAuth.

%\section{Terminology}
%\textit{Access Delegation} is a relation between two WebID resources.
%The relation describes, under which circumstances a secretary agent has access to a certain resource under the same constraints as the named agent.
%The secretary agent declares to act in the name of the named agent.

\section{Requirements}\label{sec:req}

@@

\subsection{Basic Use-Cases for Access Delegation} \label{sec:usecases}

\subsubsection{Personal Devices and Applications}

Currently, the default usage intention for WebIDs are browser-based authentication scenarios.
This is an important use-case and should be of special interest but we use the World Wide Web not only from our browsers but also from other applications and from other application on other devices such as smartphones, tablets and freedomboxes\footnote{\url{https://freedomboxfoundation.org/}}.
In a typical every day usage scenario, we use feed readers and social network clients hand in hand with web browsers.
All these applications have in common that they act in the name of its users and fetch and manipulate resources in a way that this could also be done by the user itself and with her browser.
These applications provide their service to one user only and run on a device which can be accessed by one user only (e.g. a smartphone).
This class of \textit{private applications} has more in common with the users browser and can be seen as an extension to the user itself~\cite{clark-a-1998-7-a}.

As slightly different we can identify the class of \textit{service applications}.
These applications also manipulate resources in the name of the user but they do this not only for one but for many users.
These services are hosted by a third party e.g. not by the user itself and not by the owner of the protected resource where our user has an interest in.
In addition to that, we argue that the level of trust which is given by a user is lower than to an application which runs on a personal device and can be turned off by herself.
A typical application domain for this class of applications is the Social Web more specifically the distributed and semantic Social Web, where multiple services create and manipulate artefacts like notes, comments and images and spin a data network based on a Linked Data infrastructure and shared vocabularies~\cite{tramp-s-2012--a}.

\subsubsection{Vacation Replacement}

In a typical office scenario, employees declare vacation replacements for the time of their holiday in order to achieve an uninterrupted service during that time.
%One option to allow access to the vacation replacement user is to give him the original certificate or to add his private key to the users WebID.
%This solution has two drawbacks: (1) the users certificate is burned after the vacation and she needs to create a new one, (2) a resource guard can not distinguish between the user and the vacation replacement user, which allows for abuse.
%In this context, the named agent is the employee which is on holiday and her vacation replacement is the secretary agent.

\subsubsection{Groups and Roles}
Access for \textit{working groups} to a resource is typically specified in a way that each resource guard needs to maintain (owned or cached) a list of group members of this group.
In the Social Semantic Web, these working groups have its own WebID resources which describe the group as an entity and its members as relations to their user WebIDs.
To authorize a group of agents to access a resource should be as simple as to authorize a single agent.

%In order to allow this, we define the group as the named agent and all group members as its secretary agents.
%An alternative could be to share one single group certificate with all members or to create a group certificate for each member and describe all public keys in the group WebID.
%Again, these solutions have different drawbacks: (1) a shared certificate needs to be updated each time a user leaves the group, (2) multiple group certificates for each user does not allow for distinguishing which user accessed the resource, which allows for abuse. 
%In this context, the group is the named agent and all current group members are secretary agents.\todo{A: not sure I agree (see source)}%the group is not an entity, therefore it cannot take actions by iteself. Actions are always performed by users, as part of the group.

An \textit{office role} is a construct where a user can access specific resources because of his working description.
A product manager has access to all product related resources because of her role as product manager and not because of her identity.
The role itself is identified by WebID and all access authorization rules should use this role WebID instead of a user WebID.

%In this context, the role is the named agent and the current role owner is the secretary agent.

\subsection{Implementation Requirements}

@@

\subsubsection{Distinguish secretary agents from named agents}
Delegation should not be handled by camouflaging an request of an agent in the sense that this request is indistinguishable from a named agent request.
A secretary agent should use its own certificate and it should be identified and described by its own WebID profile.
The motivation for this is twofold.
Firstly, it allows resource guards to permit or deny requests based on this information.
Secondly, applications which are secretary for many different named agents do not need to switch their certificate between requests.
Also they do not need to handle hundreds of different keys.

\subsubsection{Easy to use}
The one and only place to describe which secretary agents are allowed to operate as a named agent should be the WebID profile itself.
To grant delegated access to a secretary agent, no other actions than simply adding triples to the WebID profile should be needed.
To reset this grant the same triple can removed from the WebID profile.

\subsubsection{Linked Data and Read Write Web integration}

@@

\subsubsection{Minimal protocol footprint}

@@

\subsubsection{No enforcement to reveal agent identities}\todo{maybe we kick this ...}

@@

\section{Extending WebID for Access Delegation}\label{sec:spec}

\textit{Solution 1: Using each users WebID} vs.
\textit{Solution 2: Acting on Behalf of a User}

@@

\subsection{OAuth}
\todo{If we were to implement a secreteary like feature using OAuth what would it look like? what would the differences be?}
